---
title: "Extracting rasters"
author: "Camille Ross"
date: "8/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and load packages

```{r load_libraries, message=FALSE}
# Load libraries
library(tidyverse)
library(raster)
library(ggplot2)
library(leaflet)
library(dplyr)
library(remotes)
library(lubridate)
# Install ohwobpg package
installed <- rownames(installed.packages())
if (!("ohwobpg" %in% installed))  remotes::install_github("BigelowLab/ohwobpg", quiet = TRUE)
# Load ohwobpg package
library(ohwobpg)
```

## Initialize raster for visualization

Once the necessary packages are loaded, we will start by creating a raster object from a pre-loaded ohwobpg database.  Specifically, we will be creating a raster object of sea surface temperature (SST) data for 2018.

```{r init_raster, message=FALSE, warning=FALSE}
# Load database
path <- system.file("gom", package = "ohwobpg")
db <- read_database(path)

# Subset database
par_db <- db %>% 
  dplyr::filter(param == "sst" & 
                per == "MO" &
                dplyr::between(date, as.Date("2018-05-15"), as.Date("2018-09-26")))

# Create raster stack
sat <- par_db %>%                # start with the subset database
  as_filename(path = path) %>%   # build filenames and append to the path
  raster::stack()                # read them into a stack of images
 
# Name layers
names(sat) <- format(par_db$date, "%b")

# Read in buoy data
buoy <- read_buoy(buoy = "M01")
# Name layers
names(buoy) <- c("time", "sal", "sigma_t", "conductivity", "temp")

# Read in buoy locations
buoy_loc <- buoy_locations()[, c("lon", "lat")]

```

## Extract buoy location from satellite data

The buoy location is extracted from the satellite data raster using the code below.  For more examples of extracting point data from rasters and other raster manipulations, see https://mgimond.github.io/megug2017/#raster-manipulation-basics. 

```{r base_plot, message=FALSE, warning=FALSE}
raster::extract(sat, buoy_loc)
```

## Plot buoy location and satellite data using leaflet

Finally, we will plot the satellite data and the buoy location on a leaflet interactive map.  The leaflet::addRasterImage() function allows the user to plot raster objects on the map.  The leaflet::addCircles() function allows the user to plot point data on the map.  For a more detailed example, see https://rstudio.github.io/leaflet/raster.html.

```{r leaflet, message=FALSE, warning=FALSE}
leaflet::leaflet() %>% leaflet::addTiles(group = "Standard") %>% 
            # Add satellite imagery
            leaflet::addProviderTiles('Esri.WorldImagery', group = "Satellite") %>%
            # Define bounds
            leaflet::fitBounds(lng1 = xmin(sat), 
                               lat1 = ymin(sat), 
                               lng2 = xmax(sat), 
                               lat2 = ymax(sat)) %>%
            # Add raster data
            leaflet::addRasterImage(sat$Jul) %>%
            # Add buoy data
            leaflet::addCircles(lng = buoy_loc$lon, lat = buoy_loc$lat)
```


## Plot time series of buoy and satellite data

Finally, we plot a time series of the buoy data and satellite data.  Both datasets are formatted to plot using the ggplot2 package.

```{r time_series, message=FALSE, warning=FALSE}
# Format satellite data
sat_df <- tidyr::gather(as.data.frame(raster::extract(sat, buoy_loc)), month, value, Jun:Sep) %>%
  dplyr::mutate(month_num = 6:9)

# Format buoy data
buoy_df <- buoy %>%
  dplyr::mutate(month_num = lubridate::month(time)) %>%
  dplyr::group_by(month_num) %>%
  # Compute monthly average
  dplyr::summarise(mean = mean(temp))

# Initialize colors
colors <- c("Satellite" = "red", "Buoy" = "blue")
# Plot time series of buoy and satellite data
ggplot(data = sat_df, mapping = aes(x = month_num, y = value, color = "Satellite")) +
  geom_path() + 
  geom_path(data = buoy_df, aes(x = month_num, y = mean, color = "Buoy")) +
  labs(x = "Month",
       y = "Temp (°C)",
       color = "")
```

